Combinatorial interpretations of holonomic recurrences

### 2021-09-05 article/main.tex
Add fixups

@@

\begin{figure}[H]
    \centering
    \includesvg[scale=.7]{intro__presuf_custom}
    \caption{Visualization of prefixes and sufixes}
    \label{fig:presuf}
\end{figure}

@@

\section{Known results}%
\label{sec:known_results}

@@

Let's start with probably the most known result, full binary trees~\cite{binary}.

We define the size of a tree as a number of internal nodes (i.e.\ nodes that are not leaves). Our generating function is therefore:

@@

After a quick transformation, we get following holonomic recurrence:

\[(n + 2)B(n + 1) = 2 (2n + 1)B(n)\]

This recurrence also describes Catalan numbers.

@@

We are going to create a bijection between the left and right side of the recurrence. We will be using the term \textit{bijective language} to talk about the recurrence using the objects it describes and their transformations.

Let's translate the recurrence to the bijective language then.
\begin{itemize}
    \item \(B(n)\) is simply set of all full binary trees of size \(n\),
    \item \((n + 2) B(n + 1)\) is set of all full binary trees of size \(n + 1\) with one of the leaves pointed,
    \item \((2n + 1) B(n)\) is set of all full binary trees of size \(n\) with one of the nodes pointed (remember that the node can be internal node or leaf),
    \item \(2\) at the beginning of \(2 (2n + 1) B(n)\) means that we will be using \(2\) copies of the set defined in the previous bullet.
\end{itemize}

At the moment the left side of the recurrence describes a set of trees of size \(n + 1\), the right, on the other hand, of size \(n\). Therefore, we need to apply some transformation to make the sets equal.

Let's define two local operations: \(R_l(T, v)\) and \(R_r(T, v)\). The term \textit{local} means that the operation can operate on just some finite fragment of tree with predefined size (e.g.\ node and its 2 children, but cannot traverse the whole tree).

@@

@@

\begin{figure}[H]
    \centering
    \begin{minipage}{.25\textwidth}\includesvg[scale=0.35]{binary__remy_base}\end{minipage}%
    \(\Rightarrow\)
    \begin{minipage}{.4\textwidth}\includesvg[scale=0.5]{binary__remy_left}\end{minipage}%
    \caption{\(\enc{R_l(T, v)}\)}
    \label{fig:remy_left}
\end{figure}

\begin{figure}[H]
    \centering
    \begin{minipage}{.25\textwidth}\includesvg[scale=0.35]{binary__remy_base}\end{minipage}%
    \(\Rightarrow\)
    \begin{minipage}{.4\textwidth}\includesvg[scale=0.5]{binary__remy_right}\end{minipage}%
    \caption{\(\enc{R_r(T, v)}\)}
    \label{fig:remy_right}
\end{figure}

We can now take the bijective interpretation of \((2n + 1) B(n)\), i.e.\ trees of size \(n\) with a single node pointed. For each of such trees, \(T\) we create new one by applying \(R_l\) and \(R_r\) to the pair \(T', v\), where \(T'\) is \(T\) with removed pointers and \(v\) is a node that is pointed in \(T\). Therefore we generate \(2(2n + 1) B(n)\) trees of size \(n + 1\)

@@

@@

Let's think, what would have changed if we introduce unary nodes to the binary case~\cite{motzkin}. We will be counting edges. Note that a tree having \(n\) edges has exactly \(n + 1\) nodes.

\begin{definition}
    Unary-binary tree is such tree of which any node can have either 0, 1 or 2 children.
\end{definition}

@@

And with a bit of Maple magic~\cite{gfun}, we immediately get following recurrence:

\[(n + 2)U(n) = (2n + 1)U(n - 1) + 3(n - 1)U(n - 2)\]

The definition above is precise, but introduces a new kind of node that we would have to take care of. We can avoid it by using a trick to make the tree recall binary trees. We do not change the structure, just the notation.

\begin{definition}
    \label{def:unary_binary_2}
    Left-leaning binary tree\todo{find official name; skewed binary tree is something different} is a binary tree with the following constraint added: if a node has a right child, it also has a left child.
\end{definition}

The bijection between unary-binary and left-leaning binary trees is very simple. We just represent unary nodes as nodes with just the left child of the left-leaning binary tree. The other cases translate immediately to corresponding ones. Please note that this transformation keeps the size of the tree (i.e.~number of edges).

The next step is to make binary trees full. Let's just define virtual leaves.

\begin{definition}
    Virtual leaf is a virtual node added to the node in place of a non-existing child.
\end{definition}

As we have added \(n + 2\) virtual nodes, we have added as many virtual edges.

@@

Please note that \(n + 2\) matches the multiplier of \(U(n)\). Let's point the virtual leaves then.

To present the interpretation in bijective language, we will have to introduce several types of the the nodes (you may want to revisit Section~\ref{sub:notation} Notation):

\begin{enumerate}
    \item Vro - virtual leaf being right virtual-oneling;
    \item Vrt - virtual leaf being right virtual-twinling. We will divide this case into:
        \begin{enumerate}
            \item VrtUrv - Vri which uncle is a right virtual leaf;
            \item VrtUra - Vri which uncle is a right actual node;
            \item VrtUla - Vri which uncle is a left actual node.
        \end{enumerate}
    \item Vlt - virtual leaf left virtual-twinling. Note that from definition \ref{def:unary_binary_2}, left virtual leave cannot have a sibling being an actual node. We can divide also this case into:
        \begin{enumerate}
            \item VltUrv - Vlt which uncle is a right virtual leaf;
            \item VltUra - Vlt which uncle is a right actual node;
            \item VltUla - Vlt which uncle is a left actual node;
        \end{enumerate}
    \item I - Internal node.
\end{enumerate}

Note that it covers all possible cases of pointing virtual leaf and all the cases are mutually disjoint.

All we need to show are the transformations from the smaller cases, i.e.\ \(U(n-1)\) and \(U(n-2)\).

Let's start from \(U(n-1)\). Its multiplier is \(2n + 1\). It suggests that we can point to any node, as in a tree having \(n - 1\) edges, we have \(n\) actual and \(n + 1\) virtual ones.

In case of \(U(n-2)\), the multiplier is \(3(n-1)\), that suggests that we should perform \(3\) transformations on the trees with a pointed actual node.

@@

        &
        I \includesvg[scale=0.4]{unary_binary__1si} &
        \(\Rightarrow\)&
        Vro \includesvg[scale=0.4]{unary_binary__0fid} \\
        \hline

@@

        I \includesvg[scale=0.4]{unary_binary__2i} &
        &
        \(\Rightarrow\)&
        VrtUra \includesvg[scale=0.4]{unary_binary__0fjdOdi} \\
        \hline

@@

        I \includesvg[scale=0.4]{unary_binary__2i} &
        &
        \(\Rightarrow\)&
        VltUra \includesvg[scale=0.4]{unary_binary__0fjgOdi} \\
        \hline

        I \includesvg[scale=0.4]{unary_binary__2i} &
        &
        \(\Rightarrow\)&
        VltUla \includesvg[scale=0.4]{unary_binary__0fjgOgi} \\
        \hline
    \end{longtable}
\end{center}

As you can see, we have satisfied all of the cases of \(U(n)\) with no repetitions, which concludes the construction.

@@

\begin{definition}
    Schröder tree is a tree with no unary nodes.
\end{definition}

@@

\begin{figure}[H]
    \centering
    \includesvg[scale=0.4]{schroder__example_1}
    \includesvg[scale=0.4]{schroder__example_2}
    \includesvg[scale=0.4]{schroder__example_3}
    \includesvg[scale=0.4]{schroder__example_4}
    \(\ldots\)
    \caption{Schröder tree node}
    \label{fig:schroder_trees}
\end{figure}

This definition is quite precise, but is difficult to implement as a structural recursion. Let's transform it into something more manageable, i.e.~of finite arity.

\begin{definition}
    Well-weighted binary plane tree is a weighted tree such that:
    \begin{itemize}
        \item leaves are unweighted;
        \item internal nodes can have weight either \(1\) or \(2\);
        \item if the internal node has weight 2, then its right child is not a leaf.
    \end{itemize}
\end{definition}

We can easily define the bijection \(\Phi\) between these two models using pre-order notation:
\[\Phi(T) = \left\{\begin{array}{lcl}
            \n \no \no &:& T = \n \no \no \\
            \n\weighted{1}\; T_1\; T_2 &:& T = \n\; T_1\; T_2\\
            \n\weighted{2}\; T_1\; \Phi(\n\; T_2 \ldots T_n) &:& T = \n\; T_1\; T_2 \ldots T_n
\end{array}\right.\]

@@

Let's translate the recurrence to the bijective language then.
\begin{itemize}
    \item \((2 n - 1) T(n)\) describes well-weighted binary plane trees of size \(n\) (i.e.\ with \(n\) leaves) with some node pointed (internal or leaf),
    \item \((n + 1) T(n + 1)\) describes well-weighted binary plane trees of size \(n - 1\) with some leaf pointed,
    \item \((n - 2) T(n - 1)\) describes well-weighted binary plane trees of size \(n - 1\) with some internal node pointed.
\end{itemize}

>>>>

\theoremstyle{definition}
\newtheorem{interpretation}{Interpretation}[subsection]

@@

\begin{figure}[H]
    \centering
    \includesvg[scale=.7]{intro__presuf_custom}
    \caption{Visualization of prefixes, suffixes and subtrees}
    \label{fig:presuf}
\end{figure}

@@

As \(T\) and \(\enc{T}\) describe the same structures, notation of trees \(T\) and its encoding \(\enc{T}\) will be used interchangeably.

\section{Known results}%
\label{sec:known_results}
\todo{Add introduction}

@@

Let's start with probably the most known interpretation, full binary trees~\cite{binary}.

We define the size of a tree as the number of its internal nodes (i.e.\ nodes that are not leaves). Our generating function is therefore a solution of the following equation

@@

After a quick transformation, we get the following holonomic recurrence:

\[\left\{\begin{array}{rcl}
            b_0 &=& 1\\
            (n + 2)b_{n + 1} &=& 2 (2n + 1)b_n
\end{array}\right.\]

This equation actually have \(2\) solutions, but we can eliminate one of them due to assumption that we are counting objects, i.e.\ we are interested only in solutions that give us natural numbers.

Numbers defined as above are also known as Catalan numbers.

@@

We are going to create a bijection between the left and right side of the recurrence. We will be using the term \textit{bijective language} to talk about the recurrence using the objects it describes and their transformations.
\todo{rewrite}

\begin{interpretation}
Let's translate the recurrence to the bijective language.
\begin{itemize}
    \item \(b_n\) is simply the set of all full binary trees of size \(n\),
    \item \((n + 2) b_{n + 1}\) is the set of all full binary trees of size \(n + 1\) with one of the leaves pointed,
    \item \((2n + 1) b_{n}\) is the set of all full binary trees of size \(n\) with one of the nodes (either internal or a leaf) pointed (i.e.\ distinguished),
    \item \(2\) at the beginning of \(2 (2n + 1) b_{n}\) indicates that we will be using \(2\) copies of the set defined in the previous bullet.
\end{itemize}

At the moment the left side of the recurrence is related to trees of size \(n + 1\), the right, on the other hand, of size \(n\). Therefore, we need to apply some transformations on trees.

Let's define two local operations for tree \(T\) with a single node \(v\) pointed: \(R_l(T, v)\) and \(R_r(T, v)\). The term \textit{local} means that the operation can operate on just some finite fragment of a tree with predefined size (e.g.\ node and its 2 children, but cannot traverse the whole tree).

@@

\begin{figure}[H]
    \centering
    \begin{minipage}{.25\textwidth}\includesvg[scale=0.35]{binary__remy_base}\end{minipage}%
    \(\Rightarrow\)
    \begin{minipage}{.4\textwidth}\includesvg[scale=0.5]{binary__remy_left}\end{minipage}%
    \caption{Visualization of \(\enc{R_l(T, v)}\)}
    \label{fig:remy_left}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \begin{minipage}{.25\textwidth}\includesvg[scale=0.35]{binary__remy_base}\end{minipage}%
    \(\Rightarrow\)
    \begin{minipage}{.4\textwidth}\includesvg[scale=0.5]{binary__remy_right}\end{minipage}%
    \caption{Visualization of \(\enc{R_r(T, v)}\)}
    \label{fig:remy_right}
\end{figure}

We can now take the bijective interpretation of \((2n + 1) B(n)\), i.e.\ trees of size \(n\) with a single node pointed. For each of such trees, \(T\) we create new one by applying \(R_l\) and \(R_r\) to the pair \(T', v\), where \(T'\) is \(T\) with removed pointers and \(v\) is a node that is pointed in \(T\). Therefore we generate \(2(2n + 1) B(n)\) trees of size \(n + 1\).
\todo{rewrite}

@@

\qed
\end{interpretation}

@@

Let's think, what would have changed if we introduce unary nodes to the binary case~\cite{motzkin}.

\begin{definition}
    \label{def:unary_binary_tree}
    An unary-binary tree is an ordered tree in which any node can have either 0, 1 or 2 children.
\end{definition}

We do not allow nodes with gaps here, e.g.\ a binary node with just one child or unary node with no children.

We will define size as number of edges. Note that a tree having \(n\) edges has exactly \(n + 1\) nodes.

@@

And with a bit of Maple magic~\cite{gfun}, we immediately get the following recurrence:

\[(n + 2)u_{n} = (2n + 1)u_{n - 1} + 3(n - 1)u_{n - 2}\]

\[\left\{\begin{array}{rcl}
            u_0 &=& 1\\
            u_1 &=& 1\\
            (n + 2)u_{n} &=& (2n + 1)u_{n - 1} + 3(n - 1)u_{n - 2}
\end{array}\right.\]

Definition~\ref{def:unary_binary_tree} is precise, but introduces a new kind of node that we would have to take care of. We can avoid it by using a trick to make unary-binary trees recall binary trees. We do not change the structure, just the notation.

\begin{definition}
    \label{def:unary_binary_2}
    A left-leaning binary tree is a binary tree with the following constraint added: if a node has a right child, it also has a left child. It can have just the left child.
\end{definition}

The bijection between unary-binary trees and left-leaning binary ones is very simple. We just represent unary nodes as nodes with just a left child. The other cases translate immediately to themselves. Please note that this transformation preserves the size of the tree (i.e.~number of edges).

The next step is to make binary trees full. Let's define virtual leaves.

\begin{definition}
    Virtual node is an artificially defined node that does not affect the structure definition (i.e.\ they are not included when counting the number of children of the node).
\end{definition}

\begin{definition}
    Actual node is a node that is not virtual.
\end{definition}

\begin{definition}
    A virtual leaf is a virtual node added to the node in place of a non-existing child.
\end{definition}

As we have added \(n + 2\) virtual nodes, we have added as many virtual edges.\todo{notes?}

@@

Please note that \(n + 2\) matches the multiplier of \(u_n\) in holonomic equation. The set \((n + 2) U(n)\) corresponds to trees of size \(n\) with single virtual leaf pointed.

\begin{definition}
    We say that node \(v\) is an virtual-oneling \(\textiff\) \(v\) has no virtual siblings. Note that the root is oneling even if it has no parent.
\end{definition}

\begin{definition}
    We say that node \(v\) is virtual-twinling \(\textiff\) \(v\) has at least one virtual sibling.
\end{definition}

\begin{definition}
    If \(u\) is parent of \(v\), we say that sibling of \(u\) is an uncle of \(v\).
\end{definition}

\begin{interpretation}
To present the interpretation in bijective language, we will have to introduce several types of the nodes.

\begin{enumerate}
    \item Vro - virtual leaf being right virtual-oneling;
    \item Vrt - virtual leaf being right virtual-twinling. We will divide this case into:
        \begin{enumerate}
            \item VrtUrv - Vri which uncle is a right virtual leaf;
            \item VrtUra - Vri which uncle is a right actual node;
            \item VrtUla - Vri which uncle is a left actual node.
        \end{enumerate}
    \item Vlt - virtual leaf left virtual-twinling. Note that from definition \ref{def:unary_binary_2}, left virtual leave cannot have a sibling being an actual node. We can divide also this case into:
        \begin{enumerate}
            \item VltUrv - Vlt which uncle is a right virtual leaf;
            \item VltUra - Vlt which uncle is a right actual node;
            \item VltUla - Vlt which uncle is a left actual node;
        \end{enumerate}
    \item A - Actual node.
\end{enumerate}

Note that it covers all possible cases of nodes and all the cases are mutually disjoint. There is no such thing as VrtUlv, VltUlv nor Vlo due to Definition~\ref{def:unary_binary_2}.

All we need to show are transformations from the smaller cases, i.e.\ pointed trees from \(U(n-1)\) and \(U(n-2)\) to \(U(n)\).

Let's start from \(U(n-1)\). Its multiplier is \(2n + 1\). It suggests that we can point to any node, as in a tree having \(n - 1\) actual edges, we have \(n\) actual and \(n + 1\) virtual nodes.

In the case of \(U(n-2)\), the multiplier is \(3(n-1)\), which suggests that we should perform \(3\) distinct types of transformations on the trees with a pointed actual node.

In the table below we show transformations that are applied to trees from \(U(n-1)\) and \(U(n-2)\) in order to achieve trees from \(U(n)\). We mark pointed node with \(\pointed{}\). You can find the case name in the buttom left corners of cells.

@@

        &
        A \includesvg[scale=0.4]{unary_binary__1si} &
        \(\Rightarrow\)&
        Vro \includesvg[scale=0.4]{unary_binary__0fid} \\
        \hline

@@

        A \includesvg[scale=0.4]{unary_binary__2i} &
        &
        \(\Rightarrow\)&
        VrtUra \includesvg[scale=0.4]{unary_binary__0fjdOdi} \\
        \hline

@@

        A \includesvg[scale=0.4]{unary_binary__2i} &
        &
        \(\Rightarrow\)&
        VltUra \includesvg[scale=0.4]{unary_binary__0fjgOdi} \\
        \hline

        A \includesvg[scale=0.4]{unary_binary__2i} &
        &
        \(\Rightarrow\)&
        VltUla \includesvg[scale=0.4]{unary_binary__0fjgOgi} \\
        \hline
    \end{longtable}
\end{center}

As you can see, we have satisfied all of the cases of trees from \(U(n-2)\), \(U(n-1)\) and \(U(n)\) with no repetitions, which concludes the construction.

\qed
\end{interpretation}

@@

\begin{definition}
    \label{def:schroder_tree}
    A Schröder tree is a tree with no unary nodes. We do not allow empty children here (e.g.\ binary node with just one child).
\end{definition}

@@

\begin{figure}[H]
    \centering
    \includesvg[scale=0.4]{schroder__example_1}
    \includesvg[scale=0.4]{schroder__example_2}
    \includesvg[scale=0.4]{schroder__example_3}
    \includesvg[scale=0.4]{schroder__example_4}
    \(\ldots\)
    \caption{Possible cases of Schröder tree nodes}
    \label{fig:schroder_trees}
\end{figure}

This definition is precise, but is difficult to implement as a structural recursion. Let's define binary structure that will allow us to encode all Schröder trees.

\begin{definition}
    \label{def:well_weighted}
    A well-weighted full binary tree is a weighted tree such that:
    \begin{itemize}
        \item leaves are unweighted;
        \item internal nodes can have weight either \(1\) or \(2\);
        \item if an internal node has weight 2, then its right child is not a leaf.
    \end{itemize}
\end{definition}

Let's represent node \(v\) with a weight \(n\) as \(v\weighted{n}\).

We can easily define the bijection \(\Phi\) between models from Definition~\ref{def:schroder_tree} and~\ref{def:well_weighted} using pre-order notation:
\[\Phi(T) = \left\{\begin{array}{lcl}
            \n \no \no &:& T = \n \no \no \\
            \n\weighted{1}\; T_1\; T_2 &:& T = \n\; T_1\; T_2\\
            \n\weighted{2}\; T_1\; \Phi(\n\; T_2 \ldots T_n) &:& T = \n\; T_1\; T_2 \ldots T_n
\end{array}\right.\]
\todo{image}
\todo{continue}

@@

Let's translate the recurrence to the bijective language then.
\begin{itemize}
    \item \((2 n - 1) T(n)\) describes well-weighted binary trees of size \(n\) (i.e.\ with \(n\) leaves) with some node pointed (internal or leaf),
    \item \((n + 1) T(n + 1)\) describes well-weighted binary trees of size \(n - 1\) with some leaf pointed,
    \item \((n - 2) T(n - 1)\) describes well-weighted binary trees of size \(n - 1\) with some internal node pointed.
\end{itemize}
